On the Bit Complexity of Finding Points in Connected Components of a Smooth Real Hypersurface

### 2020-02-08 main.tex
clean up conclusion

\begin{theorem}\label{theo:main}
  Suppose that $f\in\ZZ[X_1\hdots,X_n]$ is squarefree, satisfies
  $\deg(f) \leq d, \htt(f) \leq b$, and that $V(f) \subset \C^n$ is
  smooth. Also suppose that $0 < \epsilon < 1.$

  There exists a randomized algorithm that takes $f$ and $\epsilon$
  as input and produces $n$ zero-dimensional parameterizations whose reunion includes at least one point in each connected component of
  $V(f) \cap \R^n$, with probability at least $1-\epsilon$. Otherwise, 
  the algorithm either returns a proper subset of the points, or FAIL.  In
  any case, the algorithm uses
  \[
O^{\sim}(d^{3n+1}(\log1/\epsilon)(b + \log1/\epsilon))
  \]
  bit operations. The polynomials in
  the output have degree at most $d^n,$ and height 
  \begin{align*}
  &O^{\sim}(d^{n+1}(b + \log 1 / \epsilon)). 
  \end{align*}
\end{theorem}

Here we assume that $f$ is given as a dense polynomial. However, it 
would be possible to take into account the runtime if $f$ was given 
by a {\em straight-line
  program}, that is, a sequence of operations $+,-,\times$ that takes
as input $X_1,\dots,X_n$ and evaluates $f$; the {\em length} of a
straight-line program is simply the number of operations it performs. Straigh-line programs are described in references such
as~\cite{GiHeMoPa95,GiHaHeMoMoPa97,GiHeMoMoPa98,BaGiHeMb97,EMP}. The length of the straight-line program for $f$ would
be bounded by $d^n$ and would have integer constants of height at most
$b$. Any polynomial of degree $d$ in $n$ variables can be computed by a
straight-line program of length $O\big ({{d+n} \choose n}\big)$:
evaluate all monomials of degree up to $d$ in $n$ variables, multiply
them by their respective coefficients and sum the result. However,
some inputs may be given by shorter straight-line program, and the
algorithm is able to benefit from this.

@@

In general, we cannot say much about the geometry of
$W(\pi_i,V)$, but if we apply a generic change of coordinates $\mA$,
$W(\pi_i,V)$ is then known to be equidimensional of dimension $(i-1)$
or empty~\cite{BaGiHeMb97}, and to be in so-called {\em Noether
  position}~\cite{EMP} (background notions in algebraic geometry are
in~\cite{Mumford76,Shafarevich77,ECA}; we will recall key
definitions). If this is case, it suffices to choose arbitrary
$\alpha_1,\dots,\alpha_{n-1}$ in $\Q$, and solve the systems defined
by
\begin{equation}\label{eq:syst}
X_1-\sigma_1,\dots,X_{i-1}-\sigma_{i-1}, f, \frac{\pa f}{\pa
  X_{i+1}} ,\dots,\frac{\pa f}{\pa X_n},  
\end{equation}
for $i=1,\dots,n$.  They all admit finitely many solutions, and
Theorem~2 in~\cite{EMP} proves that the union of their solution sets
contains one point on each connected component of $V \cap \R^n$.

@@

\section{Bit complexity}%
\subsection{The Algorithm}
%
For a matrix $A \in \GL(n,\Q),$ let 
\[
Y_i = X_i^{A} = [b_i] \cdot \bbm X_1\\ \vdots \\ X_n \ebm, 
\]
where $b_i$ is the i-th row of $A^{-1}$.
%
\begin{algorithm}[hbt!]
\KwIn{a strait line program $\Gamma$ that computes $f \in \Q[X_1,\hdots,X_n],$ and $d=\deg(f)$}
\KwOut{$n$ zero-dimensional parameterizations, the union of which includes at least one point in each connected component of $V(f) \cap \R^n$}
\nl Take $S := \{1,2,\hdots,3\epsilon^{-1}6n^3(2d)^{2n}\}$ and $T := \{1,2,\hdots,3\epsilon^{-1}nd^{2n}\},$ and randomly choose $A \in S^{n^2}$, and  $\sigma \in T^{n-1}$\;
\nl Apply the change of variables to obtain $f(A x)=f^A(x)$\;
\nl Use \cite{partialDerComputation} to compute the partial derivatives $\frac{\pa f^A}{\pa X_{i+1}},\hdots,\frac{\pa f^A}{\pa X_n}$\;

@@

\nl \For{$i\gets1$ \KwTo $n$}{
\nl Run \cite[Algorithm 2]{SH} $k \geq \lg(3n/\epsilon)$ times with a straight-line program computed from 
$\{Y_1-\sigma_1,\hdots,Y_{i-1}-\sigma_{i-1},f^A,\frac{\pa f^A}{\pa X_{i+1}},\hdots,\frac{\pa f^A}{\pa X_n} \}$\;
\nl Let $\mathscr{Q}_i$ be the highest cardinality zero-dimensional parameterization returned in step 5 \;
} 
\nl \Return $[\mathscr{Q}_1,\hdots,\mathscr{Q}_n]$.
\end{algorithm}
%
%
\subsection{Proof of Theorem 1.1}
%
\textbf{Bit operation cost:} The following lists the costs for each step of Algorithm 1:
%
\begin{enumerate}
    \item We defined $S = \{1,2,\hdots,2\epsilon^{-1}3n^2(2d)^{3n}\}$ and therefore the height of $a_{i,j} \in S$ is at most 
    \[
    \log 1/{\epsilon} + \log(3n^2(2d)^{2n}) = O^{\sim}(n\log d + \log 1/{\epsilon}).
    \]
    \item  We can prepare a straight-line program $\Gamma$ for $f^A$ with length at most $d^n$, and 
    by \cite[Lemma 1.2(1)(c)]{SharpEstimatesForTheEffectiveN}, height of at most 
    \[
        O^{\sim}(b + d\log 1 / \epsilon + dn).
    \]
    \item 
    After computing the partial derivatives the height grows by at most another factor of $\log d$. Furthermore, preparing the straight line program for the gradient of $f$ does not introduce large integers. The integers in $\grad (f)$ are of the same magnitude as in $\Gamma$.
    \item 
    After applying \cite[Algorithm 2]{SH} $kn$ times with $k = O(\log n + \log 1 / \epsilon)$, the total boolean cost of the algorithm is
  \[
O^{\sim}(d^{3n+1}(\log1/\epsilon)(b + \log1/\epsilon))
  \]
    where the polynomials in the output have degree at most $d^n,$ and height at most
  \[
  O^{\sim}(d^{n+1}(b + \log 1 / \epsilon)).
  \]
\end{enumerate}
%
\par \noindent 
\textbf{Probability of success:} Let $\Delta_i \in \C[\A]$ be the polynomials from Theorem 2.1. Denote by $\Delta$ the product $\Delta := \prod_{i=1}^n \D_i.$ Note that 
\begin{align}
    \deg \Delta \leq \sum_{i=1}^n \deg \Delta_i \leq 6n^3(2d)^{2n},
\end{align}
where then, if $\mA \in \C^{n \times n}$ does not cancel $\Delta,$ then  $\mA$ is invertible and $f^\mA$ satisfies $\textbf{H}_i$ for all $i \in \{1,\hdots,n\}.$ Now, assuming that $\mA$ is such a matrix, let $\Xi_i\in \C[\A]$ be the polynomials from Theorem 2.2 and denote by $\Xi$ the product $\Xi := \prod_{i=1}^n \Xi_i.$ Note that 
\begin{align}
    \deg \Xi \leq \sum_{i=1}^n \deg \Xi_i \leq nd^{2n},
\end{align}
where then, if $\sigma \in \C^{i-1}$ does not cancel $\Xi$ then $f^{\mA}$ and $\sigma$ satisfy $\textbf{H}_i^{'}$ for all $i \in \{1,\hdots,n\}.$ 
Now, by construction of \[S := \{1,2,\hdots,3\epsilon^{-1}6n^3(2d)^{2n}\}\] and \[T := \{1,2,\hdots,3\epsilon^{-1}nd^{2n}\},\] where $A \in S^{n^2}$ and  $\sigma \in T^{n-1}$ are randomly chosen, we have 
%
\[
\pr[A \in \Delta] \leq  \frac{\deg\Delta}{|S|} = \epsilon/3
\]
and
\[
\pr[\sigma \in \Xi] \leq \deg \frac{\deg\Xi}{|S|} = \epsilon/3.
\]
%
Let $\mathscr{E}$ be the event that, the parameterizations $[\mathscr{Q}_1,\hdots,\mathscr{Q}_n]$ returned in step 7 of Algorithm 1 contain all points and are correct. Then, the probability of success is equal to
\begin{align*}
 \pr[A \not\in \Delta] \times \pr[\sigma \not \in \Xi ~|~A \not\in \Delta] \times \pr[\mathscr{E}~|~A \not\in \Delta \textrm{ and } \sigma \not \in \Xi].
\end{align*}
%
Set $k = \lg(3n/\epsilon)$, with $\epsilon>0,$ so that 
\[
(1-2^{-k})^n = (1 - \epsilon/(3n))^n \geq 1 - \epsilon/3,
\]
by Bernoulli's inequality. Therefore, 
\begin{align*}
\pr[\textrm{success}] &\geq  (1- \epsilon/3)(1- \epsilon/3)\pr[\mathscr{E}~|~A \not\in \Delta \textrm{ and } \sigma \not \in \Xi]\\
&\geq (1- \epsilon/3)(1- \epsilon/3)(1-(11/32)^k)^n\\
& \geq (1- \epsilon/3)(1- \epsilon/3)(1-2^{-k})^n \\
& \geq (1- \epsilon/3)(1- \epsilon/3)(1- \epsilon/3)\\
& \geq 1 - \epsilon.
\end{align*}
%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}
%%
\end{document}

>>>>

\begin{theorem}\label{theo:main}
  Suppose that $f\in\ZZ[X_1\hdots,X_n]$ is squarefree, satisfies
  $\deg(f) \leq d$ and $\htt(f) \leq b$, and that $V(f) \subset \C^n$ is
  smooth. Also suppose that $0 < \epsilon < 1.$

  There exists a randomized algorithm that takes $f$ and $\epsilon$ as
  input and produces $n$ zero-dimensional parameterizations, the union
  of whose zeros includes at least one point in each connected
  component of $V(f) \cap \R^n$, with probability at least
  $1-\epsilon$. Otherwise, the algorithm either returns a proper
  subset of the points, or FAIL.  In any case, the algorithm uses
  \[
O^{\sim}(d^{3n+1}(\log1/\epsilon)(b + \log1/\epsilon))
  \]
  bit operations. The polynomials in
  the output have degree at most $d^n,$ and height 
  \begin{align*}
  &O^{\sim}(d^{n+1}(b + \log 1 / \epsilon)). 
  \end{align*}
\end{theorem}

Here we assume that $f$ is given as a dense polynomial.  Following
references such
as~\cite{GiHeMoPa95,GiHaHeMoMoPa97,GiHeMoMoPa98,BaGiHeMb97,EMP}, it
would be possible to refine the runtime estimate by assuming that $f$
is given by a {\em straight-line program} (that is, a sequence of
operations $+,-,\times$ that takes as input $X_1,\dots,X_n$ and
evaluates $f$). Any polynomial of degree $d$ in $n$ variables can be
computed by a straight-line program that does $O(d^n)$ operations:
evaluate all monomials of degree up to $d$ in $n$ variables, multiply
them by their respective coefficients and sum the results. However,
some inputs may be given by shorter straight-line program, and the
algorithm would actually be able to benefit from this.

@@

In general, we cannot say much about the geometry of $W(\pi_i,V)$, but
if we apply a generic change of coordinates $\mA$, $W(\pi_i,V)$ is
then known to be equidimensional of dimension $(i-1)$ or
empty~\cite{BaGiHeMb97}, and to be in so-called {\em Noether
  position}~\cite{EMP} (background notions in algebraic geometry are
in~\cite{Mumford76,Shafarevich77,ECA}; we will recall key
definitions). If this is case, it suffices to choose arbitrary
$\sigma_1,\dots,\sigma_{n-1}$ in $\Q$, and solve the systems defined
by
\begin{equation}\label{eq:syst}
X_1-\sigma_1,\dots,X_{i-1}-\sigma_{i-1}, f, \frac{\pa f}{\pa
  X_{i+1}} ,\dots,\frac{\pa f}{\pa X_n},  
\end{equation}
for $i=1,\dots,n$.  They all admit finitely many solutions, and
Theorem~2 in~\cite{EMP} proves that the union of their solution sets
contains one point on each connected component of $V \cap \R^n$.

@@

\section{Proof of the main result}

The following is our main algorithm; it expands on the sketch given in
the introduction, by quantifying the various random choices.

In step 4, we use \cite[Algorithm 2]{SH} to solve a square
system. This subroutine is randomized; in order to guarantee a higher
probability of success, we repeat the calculation $k$ times, for a
well-chosen parameter $k$.

This subroutine also requires that the input system be given by a
straight-line program. We build it (at Step 3) in the straightforward
manner already suggested in the introduction: given $f$, we can build
a straight-line program that evaluates $f$ in $O(d^n)$ operations, by
computing all monomials of degree up to $d$, multiplying them by the
corresponding coefficients in $f$, and adding results. To obtain a
straight-line program for $f^\mA$, we add $O(n^2)$ steps corresponding
to the application of the change of variables $\mA$. From this, we can
compute the required partial derivatives of $f^\mA$ for the same asymptotic
cost~\cite{BaSt83}. Finally, we add the linear equations
$X_1-\sigma_1,\hdots,X_{i-1}-\sigma_{i-1}$; this gives $\Gamma_i$.

\begin{algorithm}[!h]
\KwIn{$f \in \ZZ[X_1,\hdots,X_n]$ of degree at most $d$ and height at most $b$, and $0 < \epsilon < 1$}
\KwOut{$n$ zero-dimensional parameterizations, the union of whose zeros
  includes at least one point in each connected component of $V(f)
  \cap \R^n$}

\nl Construct $S := \{1,2,\hdots,3\epsilon^{-1}6n^3(2d)^{2n}\}$ and $T :=
\{1,2,\hdots,3\epsilon^{-1}nd^{2n}\},$ and randomly choose $\mA \in
S^{n^2}$, and $\bm \sigma \in T^{n-1}$\;

@@

\nl \For{$i\gets1$ \KwTo $n$}{
 \nl Build a straight-line program $\Gamma_i$ that computes the 
 equations
  $\{X_1-\sigma_1,\hdots,X_{i-1}-\sigma_{i-1},f^{\mA},\frac{\pa
    f^{\mA}}{\pa X_{i+1}},\hdots,\frac{\pa f^{\mA}}{\pa X_n} \}$\;

  \nl Run \cite[Algorithm 2]{SH} $k \geq \lg(3n/\epsilon)$ times
  with input $\Gamma_i$\;

  \nl Let $\mathscr{Q}_i$ be the highest cardinality
  zero-dimensional parameterization returned in step 5 \; 
} 
\nl  \Return $[\mathscr{Q}_1,\hdots,\mathscr{Q}_n]$.
\end{algorithm}

If $f^\mA$ satisfies ${\bf H}_i$, and $f^\mA$ and
$(\sigma_1,\dots,\sigma_{i-1})$ satisfy ${\bf H}'_i$ for all $i$, then
Theorem~2 in~\cite{EMP} establishes correctness.

\paragraph*{Bit operation cost:} The following lists the costs for each step of Algorithm 1:
\begin{enumerate}
    \item We defined $S = \{1,2,\hdots,2\epsilon^{-1}3n^2(2d)^{3n}\}$
      and therefore the height of any $a_{i,j} \in S$ is at most
    \[
    \log 1/{\epsilon} + \log(3n^2(2d)^{2n}) \in O^{\sim}(n\log d + \log 1/{\epsilon}).
    \]
    \item As a result, by \cite[Lemma 1.2(1)(c)]{SharpEstimatesForTheEffectiveN}, all coefficients in 
      $f^\mA$ have height $ O^{\sim}(b + d\log 1 / \epsilon + dn)$.
    \item 
    After computing the partial derivatives, the height grows by at
    most another factor of $\log d$. Thus, all polynomials in the
    system considered at Step 3 have height $ O^{\sim}(b + d\log 1 /
    \epsilon + dn)$. All integer coefficients appearing in $\Gamma_i$ 
    satisfy the same bound.    
    %% Furthermore, preparing the
    %% straight line program for the gradient of $f$ does not introduce
    %% large integers. The integers in $\grad (f)$ are of the same
    %% magnitude as in $\Gamma$.
    \item As a result, after applying \cite[Algorithm 2]{SH} $k$ times
      for each index $i$, with $k = O(\log n + \log 1 / \epsilon)$,
      the total boolean cost of the algorithm is
      \[
      O^{\sim}(d^{3n+1}(\log1/\epsilon)(b + \log1/\epsilon))
      \]
      where the polynomials in the output have degree at most $d^n,$ and height at most
      \[
      O^{\sim}(d^{n+1}(b + \log 1 / \epsilon)).
  \]
\end{enumerate}

This proves the runtime estimate, as well as our bounds on the height
of the output.

\paragraph*{Probability of success.} 
Let $\Delta_i \in \C[\A]$ be the polynomials from
Theorem~\ref{theo:gen}. Denote by $\Delta := \prod_{i=1}^n \D_i,$ and
note that
\begin{align}
    \deg \Delta \leq \sum_{i=1}^n \deg \Delta_i \leq 6n^3(2d)^{2n}.
\end{align}
If $\mA \in \C^{n \times n}$ does not cancel $\Delta,$ then $\mA$ is
invertible and $f^\mA$ satisfies $\textbf{H}_i$ for all $i \in
\{1,\hdots,n\}.$ Now, assuming that $\mA$ is such a matrix, let
$\Xi_i\in \C[\A]$ be the polynomials from Theorem 2.2 applied to
$f^{\mA}.$ Denote by $\Xi := \prod_{i=1}^n \Xi_i,$ and note that
\begin{align}
    \deg \Xi \leq \sum_{i=1}^n \deg \Xi_i \leq nd^{2n}.
\end{align}
If $\bm \sigma \in \C^{i-1}$ does not cancel $\Xi$, then $f^{\mA}$ and
$\bm \sigma$ satisfy $\textbf{H}_i^{'}$ for all $i \in
\{1,\hdots,n\};$ as we argued above, the algorithm is guaranteed to
succeed, as long as our call to Algorithm 2 in~\cite{SH} succeeds.
That latter reference establishes that by repeating the calculation 
$k$ times, and keeping the output of highest degree among those $k$
results, we succeed with probability at least $1-(1/2)^k$.

Now, by construction of
\[
S := \{1,2,\hdots,3\epsilon^{-1}6n^3(2d)^{2n}\}
\] 
and 
\[
T := \{1,2,\hdots,3\epsilon^{-1}nd^{2n}\},
\] 
where $\mA \in S^{n^2}$ and  $\bm \sigma \in T^{n-1}$ are randomly chosen, we have 
%
\[
\pr[\Delta(\mA)=0] \leq  \frac{\deg\Delta}{|S|} = \epsilon/3
\]
and
\[
\pr[\Xi(\bm \sigma)=0] \leq \deg \frac{\deg\Xi}{|T|} = \epsilon/3.
\]
%
Let $\mathscr{E}$ be the event that the parametrizations
$[\mathscr{Q}_1,\hdots,\mathscr{Q}_n]$ returned in step 6 of
Algorithm~1 are correct. Then, the probability of success is equal to
\begin{align*}
 \pr[\Delta(\mA)\ne 0] \times \pr[\Xi(\bm \sigma)\ne 0 ~|~\Delta(\mA) \ne 0] \times \pr[\mathscr{E}~|~
   \Delta(\mA)\Xi(\bm\sigma) \ne 0].
\end{align*}
%
Set $k = \lg(3n/\epsilon)$ so that 
\[
(1-2^{-k})^n = (1 - \epsilon/(3n))^n \geq 1 - \epsilon/3,
\]
by Bernoulli's inequality. Therefore, 
\begin{align*}
\pr[\textrm{success}] &\geq (1- \epsilon/3)(1-
\epsilon/3)\pr[\mathscr{E}~|~   \Delta(\mA)\Xi(\bm\sigma) \ne 0]\\ 
& \geq (1- \epsilon/3)(1-
\epsilon/3)(1-2^{-k})^n \\ & \geq (1- \epsilon/3)(1- \epsilon/3)(1-
\epsilon/3)\\ & \geq 1 - \epsilon.
\end{align*}
This finishes the proof our main theorem.

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}
%%
\end{document}
